---
title: "The PA Election Relational DB"
author: "Jonathan Tannen"
date: "February 12, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  dpi=300,
  root.dir = "C:/Users/Jonathan Tannen/Dropbox/sixty_six/posts/musa_workshop/"
)

setwd("C:/Users/Jonathan Tannen/Dropbox/sixty_six/posts/musa_workshop")
```

## Predicting the State House, Part 1
I'm hosting a workshop at Penn's [Masters of Urban Spatial Analytics](https://penniur.upenn.edu/instruction/graduate) program. In it, I'm speaking on my work [predicting the 2018 State House race](https://sixtysixwards.com/home/model-update-is-the-race-really-is-the-race-really-a-tossup-it-depends-on-the-formerly-uncontested-incumbents). It's forced me to organize my code, and I think it's useful enough that I'll be posting the course resources here as well. This tutorial is organized into four parts: the Relational Database, [Processing the GIS Geographies](02_geographies.html), [Prepping the Data for Analysis](03_rectangular_data.html), and [Making the Predictions](04_model.html). In this post, the Relational Database.

## The Relational Database

Before we can talk about modelling, I'll outline the data as I currently have it organized. I've built a "Relational Database" [(Chapter 5)](https://web.stanford.edu/~gentzkow/research/CodeAndData.pdf) out of data from the wonderful [Open Elections Project](https://github.com/openelections/openelections-data-pa). 

A relational database consists of separate rectangular datasets. Each dataset has a column or set of columns that provide a unique key. The other data in that row should be only data that applies to that row; you don't want data in multiple rows that represent the same information.

For example, part of my data-base is the data.frame `results`. The key for this data.frame is four columns: 
- `race` (e.g. \"2016 G STH STH-8\", meaning the competition for the 2016 General, State House district 8)
- `candidate` (e.g. \"JUDITH D HINES (STH)\")
- `county_code_.` and `precinct_code`, which identify unique precincts.
These four values provide unique identifiers for the row. `vote_total` is the votes received for that row. The dataframe should contain no data that isn't unique across rows. For example, it doesn't contain the candidate's gender (which should be in another data.frame, `candidates`), or the total turnout for the race (which should be in `races`). 

The magic of this setup is that it prevents duplication of data entry or other possible mistakes, while when done well, you can easily join tables to get whatever combination you need. We could join `results` with `candidates` to get `gender` if we need to for a given application.

(Note: I've already violated the structure above, since `results` contains `party`, which is duplicated across precincts and should really belong to the table `candidates_to_races`, which has unique rows for each candidate and race, ignoring precincts. Rules are made to be broken. \_shrugemoji\_)

```{r relational_db}
library(tidyverse)
load("data/relational_db.rda")
```

The database has six dataframes. `elections` has a single row for each election, meaning each year and each primary or general (separating the primaries by party). The column `election` is the unique identifier:
```{r elections_df}
head(elections)
```

`candidates` has a single row for each candidate. I've done manual cleaning to match up candidates across years who use slightly different names (e.g. different middle initials), so in theory these are unique people. This has one big exception: the same candidate running for different offices will show up as two candidates. This felt right to me, it will later mean that success in a State House race doesn't inform success is a Gubernatorial race. The table is actually unique candidate + office combinations. 

In `candidates`, column `names` is a list of all of the names that candidate has used on a ballot. Column `parties` is the list of `parties` the candidate has ever run as, and `party_guess` is my guess at whether a candidate is one of (Democrat, Republican), which I will use when they appear on the ballot as a third party. This is usually easy: a candidate who ran as a Republican four years ago might run as D/R if they're unopposed. If they've never run as a Dem or a Rep, I leave them as the third party on the ballot. The column `gender` is just my guess of the candidate's gender based on their name/googling. (Let me know if you find any issues!).  The column `candidate` is the unique identifier:

```{r candidates_df}
head(candidates)
```

Elections are divided into races. Each `race` is the competition for a given office in a given election. State-wide races are given the district `PA`. I populate the column `is_contested` as whether the second-place candidate won at least 5% of the vote; the goal is to identify whether two candidates' names were actually on the ballot, and rule out write-ins. All entries in the `.*_candidate` columns match the unique identifier in `candidates`. The column `race` is the unique identifier here:

```{r races_df}
head(races)
```

The dataframe `candidates_to_races` maps candidates to races. This includes their total `candidate_race_votes`, the `party` that they were listed on the ballot as for this race as well as `party_replaced`, which imputes Dem/Rep using `party_guess` if they're listed as a third party and there are no other Dems or no other Reps. Note: `is_incumbent` indicates whether the candidate won any seat for this office in the last cycle, regardless of district (which is important since there was redistricting in 2012.) For example, I consider Jason Altmire an incumbent in his 2012 race for USC-12, even though two years earlier he won in USC-4. 

```{r cands_to_races}
head(candidates_to_races)
```

The table `precincts_to_districts` is a mapping of which precincts voted for which districts in each year. I'll discuss this (and why we need it) in [Processing GIS](02_geographies.html).

Finally, `results` is the good stuff: the precinct-level results.
```{r results_df}
head(results)
```

That's it! Now for the hard part: [Processing the GIS geographies](02_geographies.html).